---
title: "Assignment 4: Data Wrangling"
author: "Emma Brentjens"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions
1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct7th @ 5:00pm.

## Set up your session 
1. Check your working directory, load the `tidyverse` and `lubridate` packages, and upload all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Explore the dimensions, column names, and structure of the datasets. 

```{r, message = FALSE}
#1
##Load packages
library(tidyverse)
library(tidyr)
library(lubridate)
library(dplyr)

##Determine working directory
getwd()

##upload files
O3_2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv", stringsAsFactors = T)
O3_2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv", stringsAsFactors = T)
PM25_2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv", stringsAsFactors = T)
PM25_2019 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv", stringsAsFactors = T)

#2
##Dimensions of O3_2018
str(O3_2018)
##O3_2018 is a dataframe with 20 columns and 9737 rows/observations
##column names
colnames(O3_2018)

##Dimensions of O3_2019
str(O3_2019)
##O3_2019 is a dataframe with 20 columns and 10592 rows/observations
##column names
colnames(O3_2019)

##Dimensions of PM25_2018
str(PM25_2018)
##PM25_2018 is a dataframe with 20 columns and 8983 rows/observations
##column names
colnames(PM25_2018)

##Dimensions of PM25_2019
str(PM25_2019)
##PM25_2019 is a dataframe with 20 columns and 8581 rows/observations
##column names
colnames(PM25_2019)

```

## Wrangle individual datasets to create processed files.
3. Change date to date
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3
##Determining date formats
#View(O3_2018)
#View(O3_2019)
#View(PM25_2018)
#View(PM25_2019)

##Changing class of "Date" column
O3_2018$Date <- as.Date(O3_2018$Date, format="%m/%d/%Y")
class(O3_2018$Date)

O3_2019$Date <- as.Date(O3_2019$Date, format="%m/%d/%Y")
class(O3_2019$Date)

PM25_2018$Date <- as.Date(PM25_2018$Date, format="%m/%d/%Y")
class(PM25_2018$Date)

PM25_2019$Date <- as.Date(PM25_2019$Date, format="%m/%d/%Y")
class(PM25_2019$Date)

#4
##Filtering for desired columns
O3_2018_subset <- select(O3_2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)
#View(O3_2018_subset)

O3_2019_subset <- select(O3_2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)
#View(O3_2019_subset)

PM25_2018_subset <- select(PM25_2018, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)
#View(PM25_2018_subset)

PM25_2019_subset <- select(PM25_2019, Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)
#View(PM25_2019_subset)

#5
##Editing cells in AQS_PARAMETER_DESC for PM 2.5 datasets
PM25_2018_subset$AQS_PARAMETER_DESC = "PM2.5"
PM25_2019_subset$AQS_PARAMETER_DESC = "PM2.5"

#6
##Saving processed datasets as .csv files
write.csv(O3_2018_subset, row.names = F, file = "./Data/Processed/EPAair_O3_NC2018_processed.csv")
write.csv(O3_2019_subset, row.names = F, file = "./Data/Processed/EPAair_O3_NC2019_processed.csv")
write.csv(PM25_2018_subset, row.names = F, file = "./Data/Processed/EPAair_PM25_NC2018_processed.csv")
write.csv(PM25_2019_subset, row.names = F, file = "./Data/Processed/EPAair_PM25_NC2019_processed.csv")

```


## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code. 
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels)
* Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site, aqs parameter, and county. Take the mean of the AQI value, latitude, and longitude.
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
* Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1718_Processed.csv"

```{r}
#7
##Uploading processed data
O3_2018_processed <- read.csv("./Data/Processed/EPAair_O3_NC2018_processed.csv", stringsAsFactors = T)
O3_2019_processed <- read.csv("./Data/Processed/EPAair_O3_NC2019_processed.csv", stringsAsFactors = T)
PM25_2018_processed <- read.csv("./Data/Processed/EPAair_PM25_NC2018_processed.csv", stringsAsFactors = T)
PM25_2019_processed <- read.csv("./Data/Processed/EPAair_PM25_NC2019_processed.csv", stringsAsFactors = T)

##Combining datasets
AirQuality_18_19 <- rbind(O3_2018_processed, O3_2019_processed, PM25_2018_processed, PM25_2019_processed)
#View(AirQuality_18_19)

#8
##Combined data pipe function

###intersect function?????????????

##Convert date column to date
class(AirQuality_18_19$Date)
AirQuality_18_19$Date <- as.Date(AirQuality_18_19$Date, format = "%Y-%m-%d")

AirQuality_pipe <-
  AirQuality_18_19 %>%
  filter(Site.Name == "Linville Falls" | Site.Name == "Durham Armory" | Site.Name == "Leggett" | Site.Name == "Hattie Avenue" | Site.Name == "Clemmons Middle" | Site.Name == "Mendenhall School" | Site.Name == "Frying Pan Mountain" | Site.Name == "West Johnston Co." | Site.Name == "Garinger High School" | Site.Name == "Castle Hayne" | Site.Name == "Pitt Agri. Center" | Site.Name == "Bryson City" | Site.Name == "Millbrook School") %>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  summarise(mean_AQI = mean(DAILY_AQI_VALUE),
            mean_lat = mean(SITE_LATITUDE),
            mean_lon = mean(SITE_LONGITUDE)) %>%
  mutate(Month = month(Date)) %>%
  mutate(Year = year(Date))

dim(AirQuality_pipe)

#9
##Spread dataset
AirQuality_spread <- pivot_wider(AirQuality_pipe, 
                                 names_from = AQS_PARAMETER_DESC,
                                 values_from = mean_AQI)

#View(AirQuality_spread)

#10
##Dataset dimensions
dim(AirQuality_spread) ##9 columns and 8976 rows

#11
##Saving dataset as .CSV file
write.csv(AirQuality_spread, row.names = F, file = "./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv")

```


## Generate summary tables
12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where a month and year are not available (use the function `drop_na` in your pipe). 

13. Call up the dimensions of the summary dataset.

```{r}
#12a
##Obtaining mean AQI for ozone and PM 2.5
##and remove rows with NAs for ozone and PM 2.5
mean_AQI <- 
  AirQuality_spread %>%
  drop_na(Ozone) %>%
  drop_na(PM2.5) %>%
  group_by(Site.Name, Month, Year) %>%
  summarise(Mean_AQI_O3 = mean(Ozone),
            Mean_AQI_PM25 = mean(PM2.5))

View(mean_AQI)

#12b

#13

```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: 
