---
title: "Assignment 2: Coding Basics"
author: "Emma Brentjens"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r sequence}
#1.
##creating object "seq_100": sequence from 1 to 100 increasing by 4
seq_100 <- seq(1, 100, 4) 
seq_100 ##output of sequence

#2. 
mean(seq_100) ##calculating mean of sequence
median(seq_100) ##calculating median of sequence

#3. 
##command to see if the mean is greater than the median  
isTRUE(mean(seq_100)>median(seq_100))

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r vectors}
#5. 
student_names <- c("Anne", "Brad", "Chad", "Devin") ##character vector
test_scores <- c(92, 75, 48, 83) ##numeric vector
pass_test <- c(T, T, F, T) ##logical vector 

#7.
##combining vectors into dataframe 
student_test_scores <- data.frame(student_names, test_scores, pass_test)
##creating column names for dataframe 
colnames(student_test_scores) <- c("Student Name", "Test Score", "Pass?")
##dataframe output 
student_test_scores

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A data frame can have different types of classes or modes (e.g. can combine numbers and dates), while matrices cannot.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r function}
#10.
passing_grade <- function(x){
  if(x>50){x=T} 
     else{x=F}}

anne_score <- passing_grade(92) ###not sure why output not working???????
anne_score

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: When I tried the 'ifelse' function, x=T and x=F were "unused arguments." [??????]
