---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Emma Brentjens"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r}
#1
##checking working directory
getwd()

##loading packages
library(tidyverse)
library(tidyr)
library(agricolae)
library(ggplot2)
library(lubridate)
library(dplyr)

##importing data
NTL_LTER <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

##changing dates
NTL_LTER$sampledate <- as.Date(NTL_LTER$sampledate, format = "%m/%d/%y")

##checking class of sampledate
class(NTL_LTER$sampledate)


#2
##creating theme
Emma_theme <- theme_linedraw() +
  theme(axis.text = element_text(color = "black", size = 10), 
        legend.position = "right")

##setting theme
theme_set(Emma_theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:  

> Answer:
H0: There is no significant relationship between lake temperature and depth (slope equals 0)  
Ha: Lake temperature varies significantly with water depth (slope does not equal 0)


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
##pipe function to process dataset 
NTL_LTER_Jul <- 
  NTL_LTER %>%
  mutate(month = month(sampledate)) %>%
  filter(month == "7") %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  drop_na()

#View(NTL_LTER_Jul)

#5
##water depth and temperature scatter plot
temp_depth <- ggplot(NTL_LTER_Jul, aes(x=depth, y=temperature_C)) +
  geom_point()+
  geom_smooth(method=lm) + 
  ylim(0,35) +
  xlab("Depth (m)") +
  ylab("Temperature (Celsius)")

temp_depth

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure demonstrates that temperature generally decreases with increasing water depth. The points show a greater variation in temperature at shallower depths, but temperature tapers off at around 10 meters to about 5 degrees Celsius, so the trend is not entirely linear.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
##depth and temperature linear regression
temp_depth_lm <- lm(data=NTL_LTER_Jul, temperature_C ~ depth)
summary(temp_depth_lm)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The model shows that temperature singificantly decreases with increasing water depth (p < 2e-16). The R-squared value indicates that about 73.9% of variability in temperature is due to changes in water depth. There are 9726 degrees of freedom. According to the slope of the regression line, temperature is predicted to decrease by about 1.95 degrees Celsius for every 1m increase in depth. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
##creating temperature model
temp_model <- lm(data=NTL_LTER_Jul, temperature_C ~ year4 + daynum + depth)

##stepwise AIC model selection
step(temp_model)


#10
temp_multivariate <- lm(data=NTL_LTER_Jul, temperature_C ~ year4 + daynum + depth)
summary(temp_multivariate)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The model selected by the stepwise AIC function includes year4, daynum, and depth. The R-squared value of this model is 0.7412, meaning about 74% of the variation in temperature is explained by changes in year, day of the year, and water depth. This model is improved over the depth model since the R-squared value is a bit higher. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
##July temperatures ANOVA test
temp_lake_aov <- aov(data=NTL_LTER_Jul, temperature_C ~ lakename)
summary(temp_lake_aov)

##July temperatures lm
temp_lake_lm <- lm(data=NTL_LTER_Jul, temperature_C ~ lakename)
summary(temp_lake_lm)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in temperatures between the different lakes as the p-value of the ANOVA test was less than 2e-16. The linear model, which explains about 4% of variation in temperature, shows that each lake had a significant difference in mean temperature with at least one other site. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
##depth, temperature graph
temp_depth_lake <- ggplot(NTL_LTER_Jul, aes(x=depth, y=temperature_C)) +  
  geom_point(aes(color=lakename), alpha=0.5)+
  geom_smooth(method=lm, se=F, aes(color=lakename)) +  
  ylim(0,35) +
  xlab("Depth (m)") +
  ylab("Temperature (Celsius)") +  
  guides(color=guide_legend(title="Lake Name"))

temp_depth_lake

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
##Tukey HSD test on lake mean temperatures
TukeyHSD(temp_lake_aov)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: According to the Tukey HSD test output, Paul Lake (p = 0.22) and Ward Lake (p = 0.78) have statistically similar mean temperatures as Peter Lake. There is no lake that is statistically different from every other lake. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: A two-sample T-test can be used to test if two means are significantly different. 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#18
##filtering for Crampton and Ward Lakes
NTL_LTER_Jul_Crampton_Ward <-
  NTL_LTER_Jul %>%
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake")

#View(NTL_LTER_Jul_Crampton_Ward)

##two-sample T-test
Crampton_Ward_Ttest <- t.test(data=NTL_LTER_Jul_Crampton_Ward, temperature_C ~ lakename)
Crampton_Ward_Ttest

```

>Answer: The p-value from the T-test is about 0.26 (over 0.05), so the mean temperatures at Crampton and Ward lakes are statistically similar. This p-value is different than that calculated in the TukeyHSD test (p = 0.97). According to the TukeyHSD help page, the p-value is adjusted for multiple comparisons, which likely explains this difference.
